### 2008-03-20 trunk/Papers/dl-gre-08/bisim.tex
First version of sections 2 and 3 completed.  I'm now working on the examples (section 3.2).  Perhaps sections 2 and 3 could be shortened (we are already on page 9).

\begin{theorem}[]\label{bisim}
Given a model $\gM$, if  $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
then there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
$i' \not \in |\varphi|^\gM$.
\end{theorem}

In other words, $\alc$-bisimilar elements are indistinguishable in the $\alc$ langauge.
If we only consider finite models (actually, it is enought if we consider models where each element has only a finite number of immediate successors) then the converse of Theorem~\ref{bisim} also holds.

\begin{theorem}[\cite{blac:moda01}]\label{bisim2}
Given a model $\gM$, $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
if and only if there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
$i' \not \in |\varphi|^\gM$.
\end{theorem}

By defining a suitable notion of $\el$-bisimulation, the same results can be proved for
the $\el$ language.

\cite{paig:thre87}

\cite{dovier04:_effic_algor_for_comput_bisim_equiv}

>>>>

Why are bisimulations relevant for computing referring expressions?
The following result provide us with a crucial characterization:

\begin{theorem}[\cite{blac:moda01}]\label{bisim}
Given a finite model $\gM$, $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
if and only if there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
$i' \not \in |\varphi|^\gM$.
\end{theorem}

In other words, $\alc$-bisimilar elements are indistinguishable in the $\alc$ language. Hence, bisimulation gives us a test for non-referenceability: if a
model $\gM$ contains two different individuals $i$ and $i'$ which are
$\alc$-bisimilar, then there is no formula of $\alc$ that can be used to uniquely
refer to $i$ or $i'$.  We will discuss which is the exact relation between
bisimulation and generation of referring expressions in the next section.
For the moment we just remark that a suitable notion of $\el$-bisimulation can also be defined~\cite{kurt:expr99}.

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex
First version of sections 2 and 3 completed.  I'm now working on the examples (section 3.2).  Perhaps sections 2 and 3 could be shortened (we are already on page 9).

We can connect description logics to the problem of generating
referring expressions via the concept of \emph{bisimulation classes}.
A bisimulation class is a maximal subset $C$ of the domain such that
all members of $C$ are pairwise bisimilar.  For the finite models we
deal with in GRE, this amounts to saying that there are certain
concepts that are satisfied by all individuals in $C$ and no
individuals that are not in $C$ (by Theorems~\ref{bisim} and
\ref{bisim2}).  If we represent referring expressions as DL concepts,
this means that there exists a unique referring expression for some
individual $a$ iff $a$ is alone in its bisimulation class; and if it
is, we can use one of the concepts that are characteristic for $C$ as
the referring expression.

Representing referring expressions as DL concepts is a very natural
perspective.  The purely conjunctive bag of propositional symbols that
e.g.\ \newcite{Dale1995} compute is clearly a conjunction in
description logic.  Relational expressions as in
\newcite{dale91:_gener_refer_expres_invol_relat} or
\newcite{Krahmer2003} are simply concepts of \el; for instance, ``the
book on the table'' corresponds to the concept $\mathsf{book} \sqcap
\exists \mathsf{on}. \mathsf{table}$.  And other propositional
connectives, such as disjunction and negation
\cite{deemter01:_gener_refer_expres}, can clearly be represented in
\alc\ as well.

All this means that we can reduce the problem of computing a unique RE
to the problem of computing the bisimulation classes of a model.  We
will now present two algorithms -- one for \alc\ and one for \el\ --
for doing this.  In fact, each of these algorithms computes all
bisimulation classes of a model at once, effectively computing a RE
for every single individual in the domain simultaneously.

\subsection{Bisimulation classes for \alc}

\subsection{Bisimulation classes for \el}

The algorithm we just presented is driven by case distinctions between
concepts $\varphi$ and $\neg \varphi$, and thus adds large amounts of
negations to the characteristic concepts.  This can be inconvenient
in NLG, because such concepts may be expressible only by rather clumsy
natural-language expressions. \todo{example?}  We will therefore
introduce a new algorithm which is limited to computing
(negation-free) concepts of \el.

As before, the algorithm maintains a set $\C = \{C_1,\ldots,C_n\}$ of
concepts (this time of \el) such that $\interp{C_1} \cup \ldots \cup
\interp{C_n} = \Delta$, and which it refines iteratively.  However,
where the \alc\ algorithm maintains the invariant that
$\interp{C_1},\ldots,\interp{C_n}$ is a partition of $\Delta$, we
weaken this invariant to the requirement that there are no $m \geq 2$
pairwise different indices $1 \leq i_1,\ldots,i_m \leq n$ such that
$\interp{C_{i_1}} = \interp{C_{i_2}} \cup \ldots \cup
\interp{C_{i_m}}$.  We call the concept $C_{i_1}$ \emph{subsumed} if
such a decomposition exists.

\begin{algorithm}[t]
\dontprintsemicolon
\caption{}
\label{algo:bisim-alc}
\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
\KwOut{A set \RE of formulas  such that
$\{|\varphi|^\gM \in \RE\}$ is the set of $\mathcal{L}$-bisimulation
classes of $\gM$.}

@@

\While{for some $\varphi \in \RE, |\varphi|^\gM>1$}{
   \For{$\varphi \in \RE, R \in \rel$}{
         add$_\mathcal{L}(\exists R.\varphi,\RE)$
   }
   \If{made no changes to \RE}{
      exit
      }
}

\end{algorithm}

\begin{algorithm}
\caption{add$_\alc(\varphi,\RE)$}
\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
   \If{$|\psi \sqcap \varphi|^\gM \not = \emptyset$ and
       $|\psi \sqcap \neg \varphi|^\gM \not = \emptyset$}{
         add $\psi \sqcap \varphi$ and
               $\psi \sqcap \neg \varphi$ to \RE \;
         remove $\psi$ from \RE \;
      }
   }
\end{algorithm}

\begin{algorithm}[t]
\dontprintsemicolon
\caption{add$_\el$($\varphi$, $\RE$)}
\label{algo:bisim-add-el}
\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
  \If{$\psi \sqcap \varphi$ is not subsumed in $\RE$ {\bf and}
    $|\psi \sqcap \varphi|^\gM \neq \emptyset$ {\bf and}
    $|\psi \sqcap \varphi|^\gM \neq |\psi|^\gM$}{
    add $\psi \sqcap \varphi$ to $\RE$ \;
    remove subsumed concepts from $\RE$\;
  }
}
\end{algorithm}

Algorithms~\ref{algo:bisim-el} and \ref{algo:bisim-add-el} compute
the \el\ bisimulation classes for a given model.
Algorithm~\ref{algo:bisim-el} iterates over all propositional and
relational symbols of the signature to construct new concepts until
either all concepts in $\C$ are singleton (i.e., there is only one
individual that satisfies them), or no progress has been made in the
previous iteration.  In each iteration, it calls the procedure
add$_\el$($C$, $\C$), which intersects $C$ with all current concepts
that are not already singleton, adds the result to $\C$ and then
removes all concepts in $\C$ that have become subsumed.  Removal of
subsumed concepts can be implemented efficiently by arranging the
concepts in $\C$ in a a graph that connects two concepts with an edge
if one is a maximal proper subset of the other.

@@

>>>>

We can connect description logics to the problem of generating
referring expressions via the concept of \emph{bisimulation classes}.
A bisimulation class is a maximal subset $C$ of the domain such that
all members of $C$ are pairwise bisimilar.  For the finite models we
deal with in GRE, this amounts to saying that there are certain
concepts that are satisfied by all individuals in $C$ and no
individuals that are not in $C$ (by Theorems~\ref{bisim}).  If we represent referring expressions as description logic formulas,
this means that there exists a unique referring expression for some
individual $i$ iff $i$ is alone in its bisimulation class; and if it
is, we can use one of the formulas that are characteristic for $C$ as
the referring expression.

Representing referring expressions as DL concepts is a very natural
perspective.  The purely conjunctive set of propositional symbols that,
e.g.,~\newcite{Dale1995} compute is clearly a conjunction in
description logic.  Relational expressions as in
\newcite{dale91:_gener_refer_expres_invol_relat} are concepts of \el.
We will discuss this idea further in Section~\ref{xxx}.

Summing up, we can reduce the problem of computing a unique RE
to the problem of computing the bisimulation classes of a model.  We
will now present two algorithms -- one for \alc\ and one for \el\ --
for doing this.  In fact, each of these algorithms computes all
bisimulation classes of a model at once, effectively computing a RE
for every single individual in the domain simultaneously.

\subsection{Bisimulation classes for \alc and \el}

There exists very efficient algorithms
for computing the bisimulation classes of a model~\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
These algorithms were originally used for automata minimization: given an
automata $A$ they would return a smaller automata $A'$ which recognized the
same language.  We can adapt these algorithm to compute a formula representing each bisimulation class.  Again, let's first
discuss the $\alc$ case.

We assume given a model $\gM=(\Delta^\gM, |\cdot|^\gM)$ containing all the information about the current
situation.  We will compute a set $\RE$ of $\alc$ formulas such that the
set $\{|\varphi|^\gM \mid \varphi \in \RE\}$ is a partition of $\Delta^\gM$
that contains all the bisimulation classes of $\gM$.  The pseudocode is
shown in Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.

\begin{theorem}\label{theo:bisimulation-classes}
Given a model $\gM$, the set $\mathcal{C}$ of all $\alc$-bisimulation classes
of $\gM$ is computed by Algorithms~\ref{algo:bisim-l} and~\ref{algo:bisim-add-alc} in polynomial time.  Moreover, for each class $C \in \mathcal{C}$ the algorithm returns an $\alc$ formula $\varphi$ such that
$C = |\varphi|^\gM$.
\end{theorem}

\begin{algorithm}[t]
\dontprintsemicolon
\caption{Computes $\mathcal{L}$-bisimulation classes}
\label{algo:bisim-l}
\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
\KwOut{A set \RE of formulas  such that
$\{|\varphi|^\gM \in \RE\}$ is the set of $\mathcal{L}$-bisimulation
classes of $\gM$.}

@@

\While{for some $\varphi \in \RE, |\varphi|^\gM>1$}{
   \For{$\varphi \in \RE, R \in \rel$}{
         add$_\mathcal{L}(\exists R.\varphi,\RE)$
   }
   \If{made no changes to \RE}{
      exit
      }
}
\end{algorithm}

Algorithms~\ref{algo:bisim-l} and \ref{algo:bisim-add-alc} compute
the \alc-bi\-sim\-u\-la\-tion classes for a given model.
Algorithm~\ref{algo:bisim-l} iterates over all propositional and
relational symbols of the signature to construct new formulas until
either all formulas in $\RE$ denote singletons (i.e., there is only one
individual that satisfies them), or no progress has been made in the
previous iteration.  In each iteration, it calls the procedure
add$_\alc$($\varphi$, $\RE$), which intersects $\varphi$ with any formula $\psi
\in \RE$ which does not denote a singleton and which is
not equivalent to $\varphi$ and to $\neg \varphi$. In this case it
replace $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to
$\RE$.

The $\alc$ algorithm is driven by case distinctions between
concepts $\varphi$ and $\neg \varphi$, and thus freely adds 
negations to the characteristic formulas in $\RE$.  This can be inconvenient
in natural language generation, because such concepts may be expressible only by rather clumsy natural-language expressions. We will see examples in Section~\ref{xxx}.

An algorithm which is limited to computing
(negation-free) formulas in \el can be obtained by combining
Algorithms~\ref{algo:bisim-l} and~\ref{algo:bisim-add-el}.

As before, the algorithm maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of
formulas (this time of \el) such that $\interp{\varphi_1}^\gM \cup \ldots \cup
\interp{\varphi_n}^\gM = \Delta^\gM$, and which it refines iteratively.  However,
where the \alc\ algorithm maintains the invariant that
$\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a partition of $\Delta$, we
weaken this invariant to the requirement that there are no $m \geq 2$
pairwise different indices $1 \leq i_1,\ldots,i_m \leq n$ such that
$\interp{\varphi_{i_1}} = \interp{\varphi_{i_2}} \cup \ldots \cup
\interp{\varphi_{i_m}}$.  We call the formula $\varphi_{i_1}$ \emph{subsumed} if
such a decomposition exists.

\begin{algorithm}[t]
\caption{add$_\alc(\varphi,\RE)$}
\label{algo:bisim-add-alc}
\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
   \If{$|\psi \sqcap \varphi|^\gM \not = \emptyset$ and
       $|\psi \sqcap \neg \varphi|^\gM \not = \emptyset$}{
         add $\psi \sqcap \varphi$ and
               $\psi \sqcap \neg \varphi$ to \RE \;
         remove $\psi$ from \RE \;
      }
   }
\end{algorithm}
%
\begin{algorithm}[t]
\dontprintsemicolon
\caption{add$_\el$($\varphi$, $\RE$)}
\label{algo:bisim-add-el}
\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
  \If{$\psi \sqcap \varphi$ is not subsumed in $\RE$ {\bf and}
    $|\psi \sqcap \varphi|^\gM \neq \emptyset$ {\bf and}
    $|\psi \sqcap \varphi|^\gM \neq |\psi|^\gM$}{
    add $\psi \sqcap \varphi$ to $\RE$ \;
    remove subsumed concepts from $\RE$\;
  }
}
\end{algorithm}

The procedure add$_\el$($\varphi$, $\RE$) shown in Algorithms~\ref{algo:bisim-add-el} intersects $\varphi$ with all current formulas in $\RE$
that do not denote a singleton, adds the result to $\RE$ and then
removes all formulas in $\RE$ that have become subsumed.  Removal of
subsumed formulas from $\RE$ can be implemented efficiently by arranging
them by inclusion in terms of their denotation.

@@

\subsection{Some Examples}
